\pkg{plotROC}: A Better Tool for Plotting ROC Curves

### 2015-10-15 Paper/sachs2015plotROC.tex
final responses and edits

The Receiver Operating Characteristic (ROC) curve is used to assess the
accuracy of a continuous measurement for predicting a binary outcome. In
medicine, ROC curves have a long history of use for evaluating
diagnostic tests in radiology and general diagnostics. ROC curves
originated in the field of signal detection theory .

For a continuous measurement that I denote as \(M\), convention dictates
that a test positive is defined as \(M\) equalling or exceeding some
fixed threshold \(c\): \(M \geq c\). The goal of ROC analysis is to
evaluate the classification accuracy of \(M\) in reference to the binary
outcome \(D\), which takes possible values \(0\) (negative) or \(1\)
(positive). It is implicitly assumed that the subpopulation having
\(D = 1\) tends to have larger values of \(M\) compared to the
subpopulation with \(D = 0\). The classification accuracy of \(M\) can
be evaluated by considering the confusion matrix (Table \ref{confus}).
The confusion matrix cross-classifies the predicted outcome \(M \geq c\)
versus the true outcome \(D\). The four cells of the matrix correspond
to the possible classification outcomes: a true positive, a false
positive, a true negative, and a false negative. ROC analysis assesses
the trade-offs between the test's fraction of true positives versus the
false positives.

@@

\[ FPF(c) = P\{ M \geq c | D = 0 \} \]

@@

Given a sample of test and disease status pairs,
\((M_1, D_1), \ldots, (M_n, D_n)\), I can estimate the ROC curve by
computing proportions in the diseased and healthy subgroups separately.
Specifically, given a fixed cutoff \(c\), an estimate of the \(TPF(c)\)
is

@@

where \(1\{\cdot\}\) is the indicator function. An estimate for
\(FPF(c)\) is given by a similar expression with \(D_i = 1\) replaced
with \(D_i = 0\). Calculating these proportions for \(c\) equal to each
unique value of the observed \(M_i\) yields what is known as the
empirical ROC curve estimate. The empirical estimate is a step function.
Other methods exist to estimate the ROC curve, such as the binormal
parametric estimate which can be used to get a smooth curve. There are
also extensions that allow for estimation with time-to-event outcomes
subject to censoring. For a more thorough reference on the methods and
theory surrounding ROC curves, I refer interested readers to
\citet{pepe2003statistical}.

@@

\subsection{Motivation and Literature
Review}\label{motivation-and-literature-review}

Anyone giving a cursory look at any of the major medical journals is
likely to find at least one ROC curve plot. I sought to assess the usage
of ROC curve plots and to evaluate the design choices made in the
current oncology literature by conducting a small literature review. I
searched Pubmed for clinical trials or observational studies in humans
reported in major oncology journals for the past 10 years for the terms
``ROC Curve'' OR ``ROC Analysis'' OR ``Receiver operating characteristic
curve''. The search was conducted on October 8, 2014 and returned 54
papers. From those papers, 47 images were extracted and reviewed. The
exact specifications for the Pubmed query are available in the
manuscript source files.

Each image consisted of a single ROC curve plot or a panel of multiple
plots. Each plot was inspected manually for the following design
features: the number of curves displayed, the type of axis labels
(sensitivity/ 1 - specificity or true/false positive fractions),
presence or absence of grid lines, presence or absence of a diagonal
guide line, whether any cutpoints were indicated, the type of curve
label (legend or direct label), and presence of other textual
annotations such as the AUC. The numerical results of the survey are
summarized in Table \ref{table1}.

@@

As of version 1.0.1, the \pkg{ggplot2} package \citep{ggplot2} exports
the previously internal functions \code{Geom} and \code{Stat}. This
enables developers to create their own statistical tranformations and
geometric layers, while enjoying all of the other features of
\pkg{ggplot2}. We have implemented the empirical ROC curve estimate and
the calculation of exact confidence regions as statistical
transformations: \code{stat_roc} and \code{stat_rocci}, respectively. We
have also defined geometric layers for the ROC curve and confidence
regions for the ROC curve: \code{geom_roc} and \code{geom_rocci},
respectively. For further discussion and details of the grammar of
graphics as implemented in \pkg{ggplot2}, we refer readers to
\citet{wickham2010layered} and the \pkg{ggplot2} vignettes.

To use the ROC geometric layer, I use the \code{ggplot} function to
define the aesthetic mappings, and the \code{geom_roc} function to add
an ROC curve layer. The \code{geom_roc} function requires the named
aesthetics \code{d} for disease status, and \code{m} for marker. By
default, the ROC geom and stat are linked, so that when \code{geom_roc}
is called, \code{stat_roc} does the computation, and when
\code{stat_roc} is called, \code{geom_roc} is used to plot the layer.
The disease status need not be coded as 0/1, but if it is not,
\code{stat_roc} assumes (with a warning) that the lowest value in sort
order signifies disease-free status.

@@

\begin{Schunk}
\begin{Sinput}
R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0)
R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 5, labelsize = 5, labelround = 2)
R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 50, labels = FALSE)
\end{Sinput}
\end{Schunk}

@@

This is implemented in the \code{stat_rocci} and displayed as a
\code{geom_rocci} layer. These both require the same aesthetics as the
ROC geom, \code{d} for disease status and \code{m} for marker. By
default, a set of 3 evenly spaced points along the curve are chosed to
display confidence regions. Points corresponding to the confidence
regions are distiguished from the others with a different symbol. You
can select points by passing a vector of values in the range of \code{m}
to the \code{ci.at} argument. By default, the significance level
\(\alpha\) is set to 0.05, this can be changed using the
\code{sig.level} option. An example is shown in Figure \ref{ciex}.

\begin{Schunk}
\begin{Sinput}
R> basicplot + geom_rocci()
\end{Sinput}
\begin{figure}
\includegraphics{figure/test-a-ci-1} \caption{Illustration of \pkg{plotROC} with exact confidence regions. \label{ciex}}\label{fig:test-a-ci}
\end{figure}
\begin{Sinput}
R> ## 
R> ## basicplot + geom_rocci(sig.level = .01)
R> ## ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0) +
R> ##   geom_rocci(ci.at = quantile(M1, c(.1, .4, .5, .6, .9)))
\end{Sinput}
\end{Schunk}

@@

The same objects like \code{basicplot} with Roc and/or Rocci layers can
be treated like any other ggplot objects. It can be printed to display
the figure, and other layers can be added to the plot. We provide the
function \texttt{style\_roc()} which is a layer containing a theme,
modified gridlines, and axes. Adding the \texttt{style\_roc()} layer to
the ggplot object creates a plot with sensible defaults for use in
print. This function has options for the number and location of major
and minor breaks, addition of the diagnoal guideline, axis labels, and
any theme object created for use with ggplot2 can be supplied.

@@

\begin{Schunk}
\begin{Sinput}
R> direct_label(basicplot) + style_roc()
\end{Sinput}
\begin{figure}
\includegraphics{figure/print-1} \caption{Illustration of ROC curve plot generated by \pkg{plotROC} for use in print. \label{first}}\label{fig:print}
\end{figure}
\begin{Sinput}
R> ## 
R> ## basicplot + style_roc(theme = theme_grey, xlab = "1 - Specificity")
R> ## direct_label(basicplot, labels = "Biomarker", nudge_y = -.1) + style_roc()
\end{Sinput}
\end{Schunk}

@@

Then, the dataset can be passed to the \code{ggplot} function, with the
marker name given as a grouping or faceting variable.

\begin{Schunk}
\begin{Sinput}
R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
R> ggplot(longtest, aes(d = D, m = M)) + geom_roc() + facet_wrap(~ name) + style_roc()
R> ggplot(longtest, aes(d = D, m = M, linetype = name)) + geom_roc() + geom_rocci()
R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() + style_roc()
R> pairplot <- ggplot(longtest, aes(d = D, m = M, color = name)) + 
+   geom_roc(show.legend = FALSE) + style_roc()
R> direct_label(pairplot)
R> 
R> pairplot + geom_rocci()
R> pairplot + geom_rocci(linetype = 1)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> D.cov <- rbinom(400, 1, .5)
R> gender <- c("Male", "Female")[rbinom(400, 1, .49) + 1]
R> M.diff <- rnorm(400, mean = D.cov, sd = ifelse(gender == "Male", .5, 1.5))
R> 
R> test.cov <- data.frame(D = D.cov, gender = gender, M = M.diff)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> D.ex <- test$D
R> M.ex <- test$M1
R> mu1 <- mean(M.ex[D.ex == 1])
R> mu0 <- mean(M.ex[D.ex == 0])
R> s1 <- sd(M.ex[D.ex == 1])
R> s0 <- sd(M.ex[D.ex == 0])
R> c.ex <- seq(min(M.ex), max(M.ex), length.out = 300)
R> 
R> binorm.roc <- data.frame(c = c.ex, 
+                              FPF = pnorm((mu0 - c.ex)/s0), 
+                              TPF = pnorm((mu1 - c.ex)/s1)
+                              )
R> 
R> binorm.plot <- ggplot(binorm.roc, aes(x = FPF, y = TPF, label = c)) + 
+   geom_roc(stat = "identity") + style_roc(theme = theme_grey)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> binorm.plot
\end{Sinput}
\begin{figure}
\includegraphics{figure/binormal-1} \caption[Illustration of smooth binormal ROC curve]{Illustration of smooth binormal ROC curve. \label{binorm}}\label{fig:binormal}
\end{figure}
\end{Schunk}

@@

This approach is similar to what is done in the \pkg{gridSVG}
\code{grid.animate} function, which uses the svg \code{<animate />}
tags. However, the available features were not sufficient for my needs,
which is why I used \pkg{d3.js}. There are several other \proglang{R}
packages that aim to create interactive figures. The authors of
\pkg{animint} \citep{animint} created an extensive \proglang{JavaScript}
library that creates plots in a similar way as \pkg{ggplot2}. A set of
interactive features can be added to plots using \pkg{d3.js}.
\pkg{ggvis} \citep{ggvis}, \pkg{rCharts} \citep{rcharts}, and the more
recently released \pkg{htmlwidgets} \citep{htmlwidgets} all leverage
existing charting libraries written in \proglang{JavaScript}.
\pkg{qtlcharts} \citep{qtlcharts} uses a set of custom
\proglang{JavaScript} and \pkg{d3.js} functions to visualize data from
genetic experiments. Their general approach is to manipulate the data
and create options in \proglang{R}, and then let the charting libraries
or functions handle the rendering and interactivity. \pkg{plotROC} lets
\proglang{R} do the rendering, allowing the figures to be consistent
across print and web-based media, and retaining the distinctive
\proglang{R} style. This also allows users to manipulate the figures
directly in \proglang{R} to suit their needs, using tools that are more
accessible and familiar to most \proglang{R} users.

>>>>

The Receiver Operating Characteristic (ROC) curve is used to assess the
accuracy of a continuous measurement for predicting a binary outcome. In
medicine, ROC curves have a long history of use for evaluating
diagnostic tests in radiology and general diagnostics. ROC curves
originated in the field of signal detection theory.

For a continuous measurement that I denote as \(M\), convention dictates
that a test positive is defined as \(M\) equalling or exceeding some
threshold \(c\): \(M \geq c\). The goal of ROC analysis is to evaluate
the classification accuracy of \(M\) in reference to the binary outcome
\(D\), which takes possible values \(0\) (negative) or \(1\) (positive).
It is implicitly assumed that the subpopulation having \(D = 1\) tends
to have larger values of \(M\) compared to the subpopulation with
\(D = 0\). The classification accuracy of \(M\) can be evaluated by
considering the confusion matrix (Table \ref{confus}). The confusion
matrix cross-classifies the predicted outcome \(M \geq c\) versus the
true outcome \(D\). The four cells of the matrix correspond to the
possible classification outcomes: a true positive, a false positive, a
true negative, and a false negative. ROC analysis assesses the
trade-offs between the test's fraction of true positives versus the
false positives as \(c\) varies over the range of \(M\).

@@

\[ FPF(c) = P\{ M \geq c | D = 0 \}. \]

@@

Given a sample of test and disease status pairs,
\((M_1, D_1), \ldots, (M_n, D_n)\), I can estimate the ROC curve by
computing proportions in the diseased and healthy subgroups separately.
Specifically, given a fixed cutoff \(c\), an estimate of the \(TPF(c)\)
is the proportion of test positives among the diseased subgroup:

@@

where \(1\{\cdot\}\) is the indicator function that equals 1 when the
condition inside the parentheses is true and 0 otherwise. An estimate
for \(FPF(c)\) is the proportion of test positives among the healthy
subgroup and is given by a similar expression with \(D_i = 1\) replaced
with \(D_i = 0\). Calculating these proportions for \(c\) equal to each
unique value of the observed \(M_i\) yields what is known as the
empirical ROC curve estimate. The empirical estimate is a step function.
Other methods exist to estimate the ROC curve, such as the binormal
parametric estimate which can be used to get a smooth curve. There are
also extensions that allow for estimation with time-to-event outcomes
subject to censoring. For a more thorough reference on the methods and
theory surrounding ROC curves, I refer interested readers to
\citet{pepe2003statistical}.

@@

\subsection{Literature Review}\label{literature-review}

Anyone giving a cursory look at any of the major medical journals is
likely to find at least one ROC curve plot. I sought to assess the usage
of ROC curve plots and to evaluate the design choices made in the
current oncology literature by conducting a small literature review. I
searched Pubmed for clinical trials or observational studies in humans
reported in major oncology journals for the past 10 years for the terms
``ROC Curve'' OR ``ROC Analysis'' OR ``Receiver operating characteristic
curve''. I conducted the search on October 8, 2014 and it returned 54
papers. From those papers, I extracted and reviewed 47 images. The exact
specifications for the Pubmed query are available in the manuscript
source files.

Each image consisted of a single ROC curve plot or a panel of multiple
plots. I inspected each plot manually for the following design features:
the number of curves displayed, the type of axis labels (sensitivity/ 1
- specificity or true/false positive fractions), presence or absence of
grid lines, presence or absence of a diagonal guide line, whether any
cutpoints were indicated, the type of curve label (legend or direct
label), and presence of other textual annotations such as the AUC. The
numerical results of the survey are summarized in Table \ref{table1}.

@@

As of version 1.0.1, the \pkg{ggplot2} package \citep{ggplot2} exports
the previously internal functions \code{Geom} and \code{Stat}. This
enables developers to create their own statistical tranformations
(stats) and geometric layers (geoms), while enjoying all of the other
features of \pkg{ggplot2}. We have implemented the empirical ROC curve
estimate and the calculation of exact confidence regions as statistical
transformations: \code{stat_roc} and \code{stat_rocci}, respectively. We
have also defined geometric layers for the ROC curve and confidence
regions for the ROC curve: \code{geom_roc} and \code{geom_rocci},
respectively. For further discussion and details of the grammar of
graphics as implemented in \pkg{ggplot2}, we refer readers to
\citet{wickham2010layered} and the \pkg{ggplot2} vignettes.

To use the ROC geometric layer, I use the \code{ggplot} function to
define the aesthetic mappings, and the \code{geom_roc} function to add
an ROC curve layer. The \code{geom_roc} function requires that the named
aesthetics \code{d} for disease status, and \code{m} for marker be
present in the \code{aes} function call inside \code{ggplot}. By
default, the ROC geom and stat are linked, so that when \code{geom_roc}
is called, \code{stat_roc} does the computation, and when
\code{stat_roc} is called, \code{geom_roc} is used to plot the layer.
The disease status need not be coded as 0/1, but if it is not,
\code{stat_roc} assumes (with a warning) that the lowest value in sort
order signifies disease-free status.

@@

\begin{Schunk}
\begin{Sinput}
R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0)
R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 5, 
+                                             labelsize = 5, labelround = 2)
R> ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 50, labels = FALSE)
\end{Sinput}
\end{Schunk}

@@

This is implemented in the \code{stat_rocci} and displayed as a
\code{geom_rocci} layer. These both require the same aesthetics as the
ROC geom, \code{d} for disease status and \code{m} for marker. By
default, a set of 3 evenly spaced points along the curve are chosen to
display confidence regions. Points corresponding to the confidence
regions are distiguished from the others with a different symbol. You
can select points by passing a vector of values in the range of \code{m}
to the \code{ci.at} argument. By default, the significance level
\(\alpha\) is set to 0.05, this can be changed using the
\code{sig.level} option. An example is shown in Figure \ref{ciex}.

\begin{Schunk}
\begin{Sinput}
R> basicplot + geom_rocci()
\end{Sinput}
\begin{figure}
\includegraphics{figure/test-a-ci-1} \caption{Illustration of \pkg{plotROC} with exact confidence regions. \label{ciex}}\label{fig:test-a-ci}
\end{figure}
\begin{Sinput}
R> ## basicplot + geom_rocci(sig.level = .01)
R> ## ggplot(test, aes(d = D, m = M1)) + geom_roc(n.cuts = 0) +
R> ##   geom_rocci(ci.at = quantile(M1, c(.1, .4, .5, .6, .9)))
\end{Sinput}
\end{Schunk}

@@

The same objects like \code{basicplot} with Roc and/or Rocci layers can
be treated like any other ggplot objects. It can be printed to display
the figure, and other layers can be added to the plot. We provide the
function \texttt{style\_roc()} which is a layer containing a theme,
modified gridlines, and axes. Adding the \texttt{style\_roc()} layer to
the ggplot object creates a plot with sensible defaults for use in
print. This function has options for the number and location of major
and minor breaks, addition of the diagonal guideline, axis labels, and
any theme object created for use with ggplot2 can be supplied.

@@

\begin{Schunk}
\begin{Sinput}
R> direct_label(basicplot) + style_roc()
\end{Sinput}
\begin{figure}
\includegraphics{figure/print-1} \caption{Illustration of ROC curve plot generated by \pkg{plotROC} with the styles and direct labels applied for use in print. \label{first}}\label{fig:print}
\end{figure}
\begin{Sinput}
R> ## basicplot + style_roc(theme = theme_grey, xlab = "1 - Specificity")
R> ## direct_label(basicplot, labels = "Biomarker", nudge_y = -.1)
\end{Sinput}
\end{Schunk}

@@

Then, the dataset can be passed to the \code{ggplot} function, with the
marker name given as a grouping or faceting variable. Some examples of
its usage are given below.

\begin{Schunk}
\begin{Sinput}
R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc()
R> ggplot(longtest, aes(d = D, m = M)) + geom_roc() + facet_wrap(~ name)
R> ggplot(longtest, aes(d = D, m = M, linetype = name)) + geom_roc() + geom_rocci()
R> ggplot(longtest, aes(d = D, m = M, color = name)) + geom_roc() 
R> pairplot <- ggplot(longtest, aes(d = D, m = M, color = name)) + 
+   geom_roc(show.legend = FALSE)
R> direct_label(pairplot)
R> 
R> pairplot + geom_rocci()
R> pairplot + geom_rocci(linetype = 1)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> D.cov <- rbinom(400, 1, .5)
R> gender <- c("Male", "Female")[rbinom(400, 1, .49) + 1]
R> M.diff <- rnorm(400, mean = D.cov, 
+                 sd = ifelse(gender == "Male", .5, 1.5))
R> 
R> test.cov <- data.frame(D = D.cov, gender = gender, M = M.diff)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> D.ex <- test$D
R> M.ex <- test$M1
R> mu1 <- mean(M.ex[D.ex == 1])
R> mu0 <- mean(M.ex[D.ex == 0])
R> s1 <- sd(M.ex[D.ex == 1])
R> s0 <- sd(M.ex[D.ex == 0])
R> c.ex <- seq(min(M.ex), max(M.ex), length.out = 300)
R> 
R> binorm.roc <- data.frame(c = c.ex, 
+                              FPF = pnorm((mu0 - c.ex)/s0), 
+                              TPF = pnorm((mu1 - c.ex)/s1)
+                              )
R> 
R> binorm.plot <- ggplot(binorm.roc, aes(x = FPF, y = TPF, label = c)) + 
+   geom_roc(stat = "identity") + style_roc()
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics{figure/binormal-1} \caption[Illustration of smooth binormal ROC curve]{Illustration of smooth binormal ROC curve. \label{binorm}}\label{fig:binormal}
\end{figure}
\end{Schunk}

@@

This approach is similar to what is done in the \pkg{gridSVG}
\code{grid.animate} function, which uses the svg \code{<animate />}
tags. However, the available features were not sufficient for my needs,
which is why I used \pkg{d3.js}. There are several other \proglang{R}
packages that aim to create interactive figures. The authors of
\pkg{animint} \citep{animint} created an extensive \proglang{JavaScript}
library that creates plots in a similar way as \pkg{ggplot2}. A set of
interactive features can be added to plots using \pkg{d3.js}.
\pkg{ggvis} \citep{ggvis}, \pkg{rCharts} \citep{rcharts}, and the more
recently released \pkg{htmlwidgets} \citep{htmlwidgets} all leverage
existing charting libraries written in \proglang{JavaScript}.
\pkg{qtlcharts} \citep{qtlcharts} uses a set of custom
\proglang{JavaScript} and \pkg{d3.js} functions to visualize data from
genetic experiments. Their general approach is to manipulate the data
and create options in \proglang{R}, and then let the charting libraries
or functions handle the rendering and interactivity. \pkg{plotROC} lets
\proglang{R} do the rendering, allowing the figures to be consistent
across print and web-based media, and retaining the distinctive
\proglang{R} style. This also allows users to manipulate the figures
directly in \proglang{R} to suit their needs, using tools that are more
accessible and familiar to most \proglang{R} users. Then, the
\proglang{JavaScript} adds a layer of interactivity to the rendered
figures.
